---
title: "exploration"
format: html
author: Leela Dixit, Sofia Rodas, & Vedika Shirtekar
warning: false
message: false
---

# Replicating Senko, J. 2025

This quarto document replicates the study Senko, J. 2025 studying how illuminating fish nets influences target fish catch and by-catch. This study explores the sustainable solution of using solar energy light sticks to illuminate nets with the goal of avoiding sea turtle by-catch. The original study analyzes the Bycatch Per Unit Effort (BPUE) of sea-turtles and the Catch Per Unit Effort (CPUE) for target fish species. We added an analysis of BPUE for all species to more thoroughly analyze the data.

Load libraries

```{r}
library(tidyverse)
library(mgcv)
library(stats)
library(here)
library(DHARMa)
```

Load data

```{r}
catch_data <- read_csv(here("data", "fishnet_illumination_bpue.csv"))
```

# General Linear Mixed Model (GLMM) with a Tweedie distribution

## BPUE and CPUE data frame cleaning

The paper runs a separate GLMM for Bycatch Per Unit Effort (BPUE) and Catch Per Unit Effort (CPUE). The BPUE analysis was only completed for sea turtles, not for all by-catch but we'll run both at least for exploration.

Let's make two separate data frames one for each.

BPUE is created by just selecting the appropriate rows.

```{r}
bpue <- catch_data |> 
  filter(animal_type == "total_bpue" | animal_type == "non_finfish_bpue" | animal_type == "finfish_bpue" | animal_type == "elasmobranch_bpue" | animal_type == "humboldt_squid_bpue" | animal_type == "loggerhead_biomass_bpue" | animal_type == "loggerhead_number_bpue")

turtle_bpue <- catch_data |> 
  filter(animal_type == "loggerhead_biomass_bpue" | animal_type == "loggerhead_number_bpue")
```

CPUE names are wrong. Filter for the appropriate rows, then pivot to make changing the names easier before pivotting back.

```{r}
cpue <- catch_data |> 
  filter(animal_type == "total_fish_bpue" | animal_type == "total_primary_fish_bpue" | animal_type == "halibut_bpue" | animal_type == "grouper_bpue" | animal_type == "total_secondary_fish_bpue") |> 
  pivot_wider(names_from = animal_type,
              values_from = bpue) |> 
  rename("total_fish_cpue" = "total_fish_bpue", 
         "total_primary_fish_cpue" = "total_primary_fish_bpue",
         "halibut_cpue" = "halibut_bpue",
         "grouper_cpue" = "grouper_bpue",
         "total_secondary_fish_cpue" = "total_secondary_fish_bpue") |> 
  pivot_longer(cols = c(total_fish_cpue, 
                        total_primary_fish_cpue, 
                        halibut_cpue, 
                        grouper_cpue, 
                        total_secondary_fish_cpue)) |> 
  rename("treatment"= "exp_unit",
         "catch" = "name",
         "cpue" = "value")
```

# `GLMM` on the BPUE data

### Convert appropriate variables to be a factor

```{r}
bpue <- bpue |> 
  mutate(animal_type = as.factor(animal_type), 
         treatment = as.factor(exp_unit), 
         gillnet_pair = as.factor(gillnet_pair))
```

### Initial exploration plot

```{r}
ggplot(bpue, aes(x= gillnet_pair, y = bpue, color = animal_type)) + 
  geom_point() +
  facet_wrap(~treatment) + 
  theme_minimal()

ggplot() +
  geom_histogram(data = bpue, 
                 aes(x = bpue), 
                 width = 0.2) +
  facet_wrap(~treatment) + 
  theme_minimal() +
  labs(title = "Distribution BPUE")

```

## Fit the `gam()` for the bpue

`Tweedie()` explanation. https://stat.ethz.ch/R-manual/R-devel/library/mgcv/html/Tweedie.html

```{r}
bpue_glm <- gam(bpue ~ treatment + s(gillnet_pair, bs = 're'),
                family = tw(link = log),
                data = bpue)

summary(bpue_glm)
```

# `GLMM` on the turtle BPUE data

### Convert appropriate variables to be a factor

```{r}
turtle_bpue <- turtle_bpue |> 
  mutate(animal_type = as.factor(animal_type), 
         treatment = as.factor(exp_unit), 
         gillnet_pair = as.factor(gillnet_pair))
```

### Initial exploration plot

```{r}
turtle_bpue |> 
  filter(animal_type == "loggerhead_number_bpue") |> 
ggplot(aes(x= gillnet_pair, y = bpue, color = animal_type)) + 
  geom_point() +
  facet_wrap(~treatment) + 
  theme_minimal()

turtle_bpue |> 
  filter(animal_type == "loggerhead_number_bpue") |> 
ggplot(aes(x = bpue)) +
  geom_bar(width = 0.2) +
  facet_wrap(~treatment) + 
  theme_minimal() +
  labs(title = "Distribution of Loggerhead BPUE Counts")

```

## Fit the `gam()` for the turtle bpue

```{r}
turtle_bpue_glm <- gam(bpue ~ treatment + s(gillnet_pair, bs = 're'),
                       family = tw(link = log),
                       data = turtle_bpue)
```

# `GLMM` on the CPUE data

### Convert appropriate variables to be a factor

```{r}
cpue <- cpue |> 
  mutate(catch = as.factor(catch), 
         treatment = as.factor(treatment), 
         gillnet_pair = as.factor(gillnet_pair))
```

```{r}
ggplot(cpue, aes(x= gillnet_pair, y = cpue, color = catch)) + 
  geom_point() +
  facet_wrap(~treatment) + 
  theme_minimal()

ggplot() +
  geom_histogram(data = cpue, 
                 aes(x = cpue), 
                 width = 0.2) +
  facet_wrap(~treatment) + 
  theme_minimal() +
  labs(title = "Distribution CPUE")
```

## Fit the `gam()` for the cpue

```{r}
cpue_glm <- gam(cpue ~ treatment + s(gillnet_pair, bs = 're'),
                family = tw(link = log),
                data = cpue)
```

## Calculate predictions for mean and standard errors

```{r}
# generate predictions for bpue data
prediction_turtle <- stats::predict(bpue_glm, data = bpue, se.fit = TRUE, type = "response")
# join predictions to original data
bpue_prediction <- cbind(bpue, as.data.frame(prediction_turtle))

# generate predictions for cpue data
prediction_fish <- stats::predict(cpue_glm, data = cpue, se.fit = TRUE, type = "response")
# join predictions to original data
cpue_prediction <- cbind(cpue, as.data.frame(prediction_fish))

bpue_prediction %>%
  filter(animal_type == "loggerhead_number_bpue") %>% 
  group_by(treatment) %>% 
  summarise(mean_fit = mean(fit),
            mean_se = mean(se.fit),
            .groups = "drop") %>% 
  ggplot(aes(x = treatment, y = mean_fit, fill = treatment)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_fit - mean_se, ymax = mean_fit + mean_se, width = 0.2)) +
  scale_fill_manual(values = c("control" = "#000000",
                               "illuminated" = "#39f72e")) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "#2d2f33"),
        plot.background = element_rect(fill = "#2d2f33"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        axis.title.y = element_text(color = "white"),
        legend.position = "none",
        title = element_text(color = "white")) +
  labs(x = "",
       y = "Predicted mean sea turtle BPUE\n(no. turtles per 100m net x 12h",
       title = "Predicted Sea Turtle Bycatch Decreases with Illuminated Nets")

cpue_prediction %>%
  group_by(treatment) %>% 
  summarise(mean_fit = mean(fit),
            mean_se = mean(se.fit),
            .groups = "drop") %>% 
  ggplot(aes(x = treatment, y = mean_fit, fill = treatment)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_fit - mean_se, ymax = mean_fit + mean_se, width = 0.2), color = "white") +
  scale_fill_manual(values = c("control" = "#000000",
                               "illuminated" = "#39f72e")) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "#2d2f33"),
        plot.background = element_rect(fill = "#2d2f33"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        axis.title.y = element_text(color = "white"),
        legend.position = "none",
        title = element_text(color = "white")) +
  labs(x = "",
       y = "Predicted mean target fish CPUE\n(kg fish per 100m net x 12h",
       title = "Predicted target fish CPUE increases with illuminated nets")
```

# Parametric Checks

## Distribution Determination Using AIC 

```{r}
## AIC comparison 
# Looking for lowest score to determine best distribution
bpue_gaussian <- gam(bpue ~ treatment + s(gillnet_pair, bs = 're'), 
                     family = gaussian(link = "log"),
                     data = turtle_bpue)

#Tweedie is best fit with parsimony!
AIC(turtle_bpue_glm, bpue_gaussian)
```

```{r}
## AIC comparison 
# Looking for lowest score to determine best distribution
cpue_gaussian <- gam(cpue ~ treatment + s(gillnet_pair, bs = 're'), 
                     family = gaussian(link = "log"),
                     data = cpue)

#Tweedie is best fit with parsimony!
AIC(cpue_glm, cpue_gaussian)
```

## Residual plots

In a simple linear regression (OLS), residuals can be observed directly because they are assumed to be normally distributed. In a GLMM, however, residuals are not expected to be normally distributed given that GLMMs work well for non-normally distributed data, as is common with ecological data. Here, `gam.check()` from the `mgcv` package was used to assess model fit based on residual plots, a QQ plot, and basis size checks to verify that the Tweedie distribution is appropriate for the data.

### Turtle BPUE Residual Plot Check 

```{r}
#| message: false
# Check built in residual plots
gam.check(turtle_bpue_glm)
```

`gam_check()` produces four diagnostic plots to assess the model fit:

-    QQ plot: Points follow the linear fit relatively well with some deviance in the upper tail.

-   

### CPUE Residual Plot Check

```{r}
#| message: false
# Check built in residual plots
gam.check(cpue_glm)
```

# Confidence Intervals with bootstrapping 10,000 simulations
