---
title: "Replicating Senko, J. et al. 2025"
format: html
author: Leela Dixit, Sofia Rodas, & Vedika Shirtekar
warning: false
message: false
---

# Introduction

This quarto document replicates the study Senko, J. et al. 2025 studying how illuminating fish nets influences target fish catch and by-catch. This study explores the sustainable solution of using solar energy light sticks to illuminate nets with the goal of avoiding sea turtle by-catch. The original study analyzes the Bycatch Per Unit Effort (BPUE) of sea-turtles and the Catch Per Unit Effort (CPUE) for target fish species. We added an analysis of BPUE for all species to more thoroughly analyze the data.

Load libraries

```{r}
library(tidyverse)
library(mgcv)
library(stats)
library(here)
library(DHARMa)
library(kableExtra)
```

Load data

```{r}
catch_data <- read_csv(here("data", "fishnet_illumination_bpue.csv"))
```

## Data Description

The data for this study, Senko et al 2025, consists of 4 columns and 784 rows.

| Column Header | gillnet_pair | animal_type | exp_unit | bpue |
|---------------|---------------|---------------|---------------|---------------|
| Description | A numeric value 1-28 indicating which gillnet pair this sample is from. | Categorical variable describing which animal type this observation is for. | Categorical variable indicating treatment (illuminated) or control. | Numeric value of catch. This number is on the scale of bpue or cpue depending on animal type. |

Our key outcome variable for this study is bpue/cpue, with variations of animal type as a grouping variable (turtle bpue or target fish cpue). Our treatment is exp_unit, indicating if that gillnet bycatch observation was from an illuminated net or not. This study compared 28 gillnet pairs of equal size/material over a time period of July-September 2019 (assessing sea turtle bycatch) and February-March 2020 (assessing target fish catch). These experiments were conducted in Baja California Sur (BCS), Mexico, where bottom-set gillnet results in high bycatch rates and subsequent mortality of sea turtles.

Each row is an observation of total bpue/cpue for that gillnet pair, treatment or control type, and animal type for the time period of this experiment. We have cleaned column headers/variable names and have pivoted the data from wide to long before reading in data to this script. We did not perform any data cleaning that resulted in loss of data.

# General Linear Mixed Model (GLMM) with a Tweedie distribution

## BPUE and CPUE data frame cleaning

The paper runs a separate GLMM for Bycatch Per Unit Effort (BPUE) and Catch Per Unit Effort (CPUE). The BPUE analysis was only completed for sea turtles, not for all by-catch but we'll run both at least for exploration.

Let's make two separate data frames one for each.

BPUE is created by just selecting the appropriate rows.

```{r}
bpue <- catch_data |> 
  filter(animal_type == "total_bpue" | animal_type == "non_finfish_bpue" | animal_type == "finfish_bpue" | animal_type == "elasmobranch_bpue" | animal_type == "humboldt_squid_bpue" | animal_type == "loggerhead_biomass_bpue" | animal_type == "loggerhead_number_bpue")

turtle_bpue <- catch_data |> 
  filter(animal_type == "loggerhead_biomass_bpue" | animal_type == "loggerhead_number_bpue")
```

CPUE names are wrong. Filter for the appropriate rows, then pivot to make changing the names easier before pivotting back.

```{r}
cpue <- catch_data |> 
  filter(animal_type == "total_fish_bpue" | animal_type == "total_primary_fish_bpue" | animal_type == "halibut_bpue" | animal_type == "grouper_bpue" | animal_type == "total_secondary_fish_bpue") |> 
  pivot_wider(names_from = animal_type,
              values_from = bpue) |> 
  rename("total_fish_cpue" = "total_fish_bpue", 
         "total_primary_fish_cpue" = "total_primary_fish_bpue",
         "halibut_cpue" = "halibut_bpue",
         "grouper_cpue" = "grouper_bpue",
         "total_secondary_fish_cpue" = "total_secondary_fish_bpue") |> 
  pivot_longer(cols = c(total_fish_cpue, 
                        total_primary_fish_cpue, 
                        halibut_cpue, 
                        grouper_cpue, 
                        total_secondary_fish_cpue)) |> 
  rename("treatment"= "exp_unit",
         "catch" = "name",
         "cpue" = "value")
```

# `GLMM` on the BPUE data

### Convert appropriate variables to be a factor

```{r}
bpue <- bpue |> 
  mutate(animal_type = as.factor(animal_type), 
         treatment = as.factor(exp_unit), 
         gillnet_pair = as.factor(gillnet_pair))
```

### Initial exploration plot

```{r}
ggplot() +
  geom_histogram(data = bpue, 
                 aes(x = bpue,
                     fill = treatment), 
                 width = 0.2) +
  facet_wrap(~treatment) + 
  theme_minimal() +
  scale_fill_manual(values = c("control" = "#000000",
                               "illuminated" = "#39f72e")) +
  labs(title = "Distribution BPUE") +
  theme(panel.background = element_rect(fill = "#2d2f33"),
        plot.background = element_rect(fill = "#2d2f33"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        axis.title.y = element_text(color = "white"),
        legend.position = "none",
        title = element_text(color = "white"),
        strip.text = element_text(color = "white")) 
```

## Fit the `gam()` for the bpue

`Tweedie()` explanation. https://stat.ethz.ch/R-manual/R-devel/library/mgcv/html/Tweedie.html

```{r}
bpue_glm <- gam(bpue ~ treatment + s(gillnet_pair, bs = 're'),
                family = tw(link = log),
                data = bpue)

summary(bpue_glm)
```

# `GLMM` on the turtle BPUE data

### Convert appropriate variables to be a factor

```{r}
turtle_bpue <- turtle_bpue |> 
  mutate(animal_type = as.factor(animal_type), 
         treatment = as.factor(exp_unit), 
         gillnet_pair = as.factor(gillnet_pair))
```

### Initial exploration plot

```{r}
turtle_bpue |> 
  filter(animal_type == "loggerhead_number_bpue") |> 
ggplot(aes(x = bpue,
           fill = treatment)) +
  geom_bar(width = 0.2) +
  facet_wrap(~treatment) + 
    scale_fill_manual(values = c("control" = "#000000",
                               "illuminated" = "#39f72e")) +
  theme_minimal() +
  labs(title = "Distribution of Loggerhead BPUE Counts") +
theme(panel.background = element_rect(fill = "#2d2f33"),
        plot.background = element_rect(fill = "#2d2f33"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        axis.title.y = element_text(color = "white"),
        legend.position = "none",
        title = element_text(color = "white"),
        strip.text = element_text(color = "white")) 

```

## Fit the `gam()` for the turtle bpue

```{r}
turtle_bpue_glm <- gam(bpue ~ treatment + s(gillnet_pair, bs = 're'),
                       family = tw(link = log),
                       data = turtle_bpue)
```

# `GLMM` on the CPUE data

### Convert appropriate variables to be a factor

```{r}
cpue <- cpue |> 
  mutate(catch = as.factor(catch), 
         treatment = as.factor(treatment), 
         gillnet_pair = as.factor(gillnet_pair))
```

```{r}
ggplot() +
  geom_histogram(data = cpue, 
                 aes(x = cpue,
                     fill = treatment), 
                 width = 0.2) +
  facet_wrap(~treatment) + 
    scale_fill_manual(values = c("control" = "#000000",
                               "illuminated" = "#39f72e")) +
  theme_minimal() +
  labs(title = "Distribution CPUE") +
theme(panel.background = element_rect(fill = "#2d2f33"),
        plot.background = element_rect(fill = "#2d2f33"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        axis.title.y = element_text(color = "white"),
        legend.position = "none",
        title = element_text(color = "white"),
        strip.text = element_text(color = "white")) 
```

## Fit the `gam()` for the cpue

```{r}
cpue_glm <- gam(cpue ~ treatment + s(gillnet_pair, bs = 're'),
                family = tw(link = log),
                data = cpue)
```

## Calculate predictions for mean and standard errors

```{r}
# generate predictions for bpue data
prediction_turtle <- stats::predict(bpue_glm, 
                                    data = bpue, 
                                    se.fit = TRUE, 
                                    type = "response")
# join predictions to original data
bpue_prediction <- cbind(bpue, as.data.frame(prediction_turtle))

# generate predictions for cpue data
prediction_fish <- stats::predict(cpue_glm, 
                                  data = cpue, 
                                  se.fit = TRUE, 
                                  type = "response")
# join predictions to original data
cpue_prediction <- cbind(cpue, as.data.frame(prediction_fish))

bpue_prediction %>%
  filter(animal_type == "loggerhead_number_bpue") %>% 
  group_by(treatment) %>% 
  summarise(mean_fit = mean(fit),
            mean_se = mean(se.fit),
            .groups = "drop") %>% 
  ggplot(aes(x = treatment, y = mean_fit, fill = treatment)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_fit - mean_se, ymax = mean_fit + mean_se, width = 0.2)) +
  scale_fill_manual(values = c("control" = "#000000",
                               "illuminated" = "#39f72e")) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "#2d2f33"),
        plot.background = element_rect(fill = "#2d2f33"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        axis.title.y = element_text(color = "white"),
        legend.position = "none",
        title = element_text(color = "white")) +
  labs(x = "",
       y = "Predicted mean sea turtle BPUE\n(no. turtles per 100m net x 12h",
       title = "Predicted Sea Turtle Bycatch Decreases with Illuminated Nets")

cpue_prediction %>%
  group_by(treatment) %>% 
  summarise(mean_fit = mean(fit),
            mean_se = mean(se.fit),
            .groups = "drop") %>% 
  ggplot(aes(x = treatment, y = mean_fit, fill = treatment)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_fit - mean_se, ymax = mean_fit + mean_se, width = 0.2), color = "white") +
  scale_fill_manual(values = c("control" = "#000000",
                               "illuminated" = "#39f72e")) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "#2d2f33"),
        plot.background = element_rect(fill = "#2d2f33"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        axis.title.y = element_text(color = "white"),
        legend.position = "none",
        title = element_text(color = "white")) +
  labs(x = "",
       y = "Predicted mean target fish CPUE\n(kg fish per 100m net x 12h",
       title = "Predicted target fish CPUE increases with illuminated nets")
```

# Parametric and Assumption Checks

## Distribution Determination Using AIC

AIC (Akaike Information Criterion) was used to compare model fit across different distributions while accounting for model complexity. Lower AIC values indicate a better balance between goodness of fit and simplicity. The Tweedie distribution had the lowest AIC for both models, indicating it provided the best fit to the data.

```{r}
## AIC comparison 
# Looking for lowest score to determine best distribution
bpue_gaussian <- gam(bpue ~ treatment + s(gillnet_pair, bs = 're'), 
                     family = gaussian(link = "log"),
                     data = turtle_bpue)

# Tweedie is best fit with parsimony
AIC(turtle_bpue_glm, bpue_gaussian)
```

```{r}
## AIC comparison 
# Looking for lowest score to determine best distribution
cpue_gaussian <- gam(cpue ~ treatment + s(gillnet_pair, bs = 're'), 
                     family = gaussian(link = "log"),
                     data = cpue)

# Tweedie is best fit with parsimony
AIC(cpue_glm, cpue_gaussian)
```

## Residual Plots

In a simple linear regression (OLS), residuals can be observed directly because they are assumed to be normally distributed. In a GLMM, however, residuals are not expected to be normally distributed given that GLMMs work well for non-normally distributed data, as is common with ecological data. Here, residual plots were created manually to assess model fit and verify that the Tweedie distribution with a log link is appropriate is for the data. The residual plot checks for systematic patterns that would indicate a poor fit.

```{r}
#| message: false
## Create residuals plot based on fitted values and residuals of glmm

data.frame(
  fitted = fitted(turtle_bpue_glm), # Use fitted() to simulate glmm fitted values 
  residuals = residuals(turtle_bpue_glm)) %>% # Use residuals() to obtain glmm fitted values 
  
  # Plot
  ggplot(aes(x = fitted, y = residuals)) + geom_point(shape = 21) + 
  geom_hline(yintercept = 0, color = "red", linetype = "solid") + 
  theme_minimal() + labs(x = "Fitted Values",
         y = "Residuals",
         title = "Residuals vs Fitted of Turtle BPUE GLMM")
```

```{r}
#| message: false
## Create residuals plot based on fitted values and residuals of glmm

data.frame(
  fitted = fitted(cpue_glm), # Use fitted() to simulate glmm fitted values 
  residuals = residuals(cpue_glm)) %>% # Use residuals() to obtain glmm fitted values 
  
  # Plot
  ggplot(aes(x = fitted, y = residuals)) + geom_point(shape = 21) + 
  geom_hline(yintercept = 0, color = "red", linetype = "solid") + 
  theme_minimal() + labs(x = "Fitted Values",
         y = "Residuals",
         title = "Residuals vs Fitted of CPUE GLMM")
```

## Log Link Assumption Check

Here, the log-transformed sea turtle BPUE and total fish catch (CPUE) outcomes were plotted against treatment to verify the multiplicative (non-fixed) treatment effect assumed by the log link.

```{r}
#| warning: false
# Turtle BPUE: Predictor vs log(outcome)
turtle_bpue <- turtle_bpue %>% 
  mutate(log_bpue = log(bpue))

# Plot
turtle_bpue %>% ggplot(aes(x = treatment, y = log_bpue, 
                           color = treatment)) +
  geom_jitter(width = .2, alpha = .6) +
  geom_boxplot(aes(fill = treatment), alpha = .3) +
  scale_color_manual(values = c("control" = "#000000", "illuminated" = "#39f72e")) + 
  scale_fill_manual(values = c("control" = "#000000", "illuminated" = "#39f72e")) +
  theme_minimal() +  
  labs(x = "Treatment", y = "log(Sea Turtle BPUE)", title = "Treatment vs log(BPUE)")+ 
  guides(fill = "none") +
  theme(panel.background = element_rect(fill = "#2d2f33"),
        plot.background = element_rect(fill = "#2d2f33"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        axis.title.y = element_text(color = "white"),
        legend.position = "none",
        title = element_text(color = "white")) 
```

```{r}
#| warning: false
# Total Fish Catch CPUE: Predictor vs log(outcome)
cpue <- cpue %>% 
  mutate(log_cpue = log(cpue))

# Plot
cpue %>% ggplot(aes(x = treatment, y = log_cpue, 
                           color = treatment)) +
  geom_jitter(width = .2, alpha = .6) +
  geom_boxplot(aes(fill = treatment), alpha = .3) +
  scale_color_manual(values = c("control" = "#000000", "illuminated" = "#39f72e")) + 
  scale_fill_manual(values = c("control" = "#000000", "illuminated" = "#39f72e")) +
  theme_minimal() +  
  labs(x = "Treatment", y = "log(CPUE)", title = "Treatment vs log(CPUE)")+ 
  guides(fill = "none") +
  theme(panel.background = element_rect(fill = "#2d2f33"),
        plot.background = element_rect(fill = "#2d2f33"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        axis.title.y = element_text(color = "white"),
        legend.position = "none",
        title = element_text(color = "white")) 
```

Among non-zero observations, the control group shows consistenly higher log(BPUE) values than the illuminated group. This helps support the assumption that the illumination reduces bycatch proportionally rather than by a fixed amount.

# Confidence Intervals

### Turtle by-catch confidence intervals

Senko, J. et al. 2025 state they utilize a bootstrap analysis with 10,000 simulations to calculate confidence intervals. Bootstrapping creates confidence intervals by creating simulations through resampling the data with replacement to analyze an estimate of the true populations' data distribution. To complete bootstrapping six steps are taken:

1.  Create predictions for the outcome based on the sample data.

2.  Sample from the predictions with replacement. This is repeated a defined number of times. To maintain consistency with the predictions, sampling with replacement is completed the based on the number of predicted observations.

3.  Run a GLMM Tweedie distribution analysis on the boot sample.

4.  Save the prediction values

5.  Repeat the bootstrapping practice to gat a better estimate. In this case the bootstrap was run 10,000 times.

6.  Calculate the desired statistic from these bootstrapped estimates of the outcome. In this case, confidence intervals are calculated.

The above steps are completed for each set of data for which the confidence intervals need to be calculated. These steps were run for the turtle by-catch data and the catch per unit effort data.

# Bootstrap 10,000 simulations

Note: Each bootstrap runs with a warning since not all of the levels of `gillnet_pair`, which is classified as a factor, are included in each iteration of the bootstraps and therefore predictions for that factor cannot be calculated. After much research this does not affect the effectiveness of the bootstrap. These warning are suppressed.

### Run bootstrap for turtle by-catch

Create predictions for only turtle by-catch.

```{r}
# generate predictions for bpue data
prediction_turtle <- predict(turtle_bpue_glm, 
                             data = turtle_bpue, 
                             se.fit = TRUE, 
                             type = "response")

# join predictions to original data
turtle_prediction <- cbind(turtle_bpue, as.data.frame(prediction_turtle))
```

Bootstrap the turtle by-catch data.

```{r}
#| warning: false

# reiterate through the boot 1000 times
boot <-  map_dfr(1:10000, \(i) {

  # sample with replacement
  boot_sample <- slice_sample(turtle_bpue, n=112, replace = TRUE)
  
  # run a GLMM on every single boot sample
  boot_mod <- gam(bpue ~ treatment + s(gillnet_pair, bs = 're'),
                       family = tw(link = "log"),
                       data = boot_sample)

  # predict the boots values
  mutate(turtle_prediction, 
         turtle_pred = predict(boot_mod, 
                 newdata = turtle_prediction, 
                 type = "response"),
         boot = i)
})
```

Create confidence intervals utilizing the bootstrapped data.

```{r}
turtle_boot_ci <- boot |> 
  summarise(bpue_lwr = quantile(turtle_pred, 0.025),
            bpue_upr = quantile(turtle_pred, 0.975),
            .by = c(animal_type, gillnet_pair, treatment))
```

### Run bootstrap for CPUE

```{r}
#| warning: false

# reiterate through the boot 1000 times
boot <-  map_dfr(1:10000, \(i) {

  # sample with replacement
  boot_sample <- slice_sample(cpue, n=280, replace = TRUE)
  
  # run a GLMM on every single boot sample
  boot_mod <- gam(cpue ~ treatment + s(gillnet_pair, bs = 're'),
                     family = tw(link = "log"),
                     data = boot_sample)

    # predict the boots values 
  mutate(cpue_prediction, 
         cpue_pred = predict(boot_mod, 
                 newdata = cpue_prediction, 
                 type = "response"),
         boot = i)
})
```

Create confidence intervals utilizing the bootstrapped data.

```{r}
cpue_boot_ci <- boot |>
  summarise(cpue_lwr = quantile(cpue_pred, 0.025),
            cpue_upr = quantile(cpue_pred, 0.975),
            .by = c(catch, gillnet_pair, treatment))
```

```{r}
#| echo: false

# lower boot CI
turtle_boot_ci <- turtle_boot_ci |> 
  group_by(treatment) |> 
  mutate(lwr_avg = mean(bpue_lwr, na.rm = TRUE))

# upper boot CI
turtle_boot_ci <- turtle_boot_ci |> 
  group_by(treatment) |> 
  mutate(upr_avg = mean(bpue_upr, na.rm = TRUE))

ill_turtle <- turtle_boot_ci |> 
  filter(treatment == "illuminated") |> 
  head(n = 1)

cont_turtle <- turtle_boot_ci |> 
  filter(treatment == "control") |> 
  head(n = 1)

CI_df <- tibble(
  Treatment = c("Illuminated", "Control"),
  LowerCI = c(ill_turtle$lwr_avg, cont_turtle$lwr_avg),
  UpperCI = c(ill_turtle$upr_avg, cont_turtle$upr_avg)
)

turtle_CI <- CI_df |> 
  kable() |> 
  kable_styling()

turtle_CI |> 
  add_header_above(c("Turtle BPUE Confidence Intervals" = 3))

```

```{r}
#| echo: false

# lower boot CI
cpue_boot_ci <- cpue_boot_ci |> 
  group_by(treatment) |> 
  mutate(lwr_avg = mean(cpue_lwr, na.rm = TRUE))

# upper boot CI
cpue_boot_ci <- cpue_boot_ci |> 
  group_by(treatment) |> 
  mutate(upr_avg = mean(cpue_upr, na.rm = TRUE))

ill_cpue <- cpue_boot_ci |> 
  filter(treatment == "illuminated") |> 
  head(n = 1)

cont_cpue <- cpue_boot_ci |> 
  filter(treatment == "control") |> 
  head(n = 1)

CI_df <- tibble(
  Treatment = c("Illuminated", "Control"),
  LowerCI = c(ill_cpue$lwr_avg, cont_cpue$lwr_avg),
  UpperCI = c(ill_cpue$upr_avg, cont_cpue$upr_avg)
)

cpue_CI <- CI_df |> 
  kable() |> 
  kable_styling()

cpue_CI |> 
  add_header_above(c("CPUE Confidence Intervals" = 3))

```

These confidence intervals shows where we think the true population mean falls based on our predictions. The confidence interval for the mean turtle BPUE is significantly smaller in the illuminated gillnet pairs than the confidence interval for the mean turtle BPUE in the control gillnet pairs. The confidence intervals for the mean CPUE are approximately the same both the illuminated and control groups. This shows that illumination has a greater effect on the mean of turtle BPUE.

# Conclusions

-   Causal identification credible?

    -   Some evidence for a credible causal identification (see assumptions below), but also some issues:

        -   bycatch and target fish experiments done during different seasons

        -   between-sample conditions were not kept the same

        -   target fish experiment had a much smaller sample size compared to sea turtle bycatch sample size

-   assumptions what are they, did they have them?

    -   Response variables followed Tweedie exponential dispersion model (variance is proportional to power of the mean `p`)
    -   Data are non-negative, continuous (skewed right)
    -   The relationship between predictor and log-transformed mean is linear
    -   quantile residuals follow a normal distribution
    -   power law: spread of residuals gets larger as the mean gets larger

-   sampling validity who is in the sample, can results be generalized?

    -   very small sample size of target fish (n = 21) vs sea turtle bycatch (n = 67)
    -   only one sampling location

-   measurement are outcome/treatment variables measured accurately without bias?

    -   Sampling for sea turtle bycatch was checked every hour, but target fish experiment was not. All deployed for the same amount of time though (overnight 11-12 hours).
    -   28 net pairs for sea turtle bycatch, and 14 net pairs for target fish experiments.
    -   gillnet pairs were identical and similar material to what is already used, and deployed during times fishing would naturally occur.

-   other limitations

    -   The authors were looking only at blinking lights as illumination and comparing to previous studies looking only at solid light illumination.

# References

-   Senko, J. F., J. Wang, K. M. Burgher, et al. 2025. “Harnessing SolarEnergy to Reduce Sea Turtle Bycatch.” Conservation Letters 18, no. 5:e13151.
