---
title: "exploration"
format: html
warning: false
message: false
---

Load libraries
```{r}
library(tidyverse)
library(mgcv)
library(stats)
library(here)
```

Load data
```{r}
catch_data <- read_csv(here("data", "fishnet_illumination_bpue.csv"))
# biomass <- read_csv(here("data", "fishnet_illumination_biomass.csv"))
```
# General Linear Mixed Model (GLMM) with a Tweedie distribution

The paper runs a separate GLMM for Bycatch Per Unit Effort (BPUE) and Catch Per Unit Effort (CPUE). The BPUE analysis was only completed for sea turtles, not for all by-catch but we'll run both at least for exploration.

Let's make two separate data frames one for each.

BPUE is created by just selecting the appropriate rows. 
```{r}
bpue <- catch_data |> 
  filter(animal_type == "total_bpue" | animal_type == "non_finfish_bpue" | animal_type == "finfish_bpue" | animal_type == "elasmobranch_bpue" | animal_type == "humboldt_squid_bpue" | animal_type == "loggerhead_biomass_bpue" | animal_type == "loggerhead_number_bpue")
```

CPUE names are wrong. Filter for the appropriate rows, then pivot to make changing the names easier before pivotting back. 
```{r}
cpue <- catch_data |> 
  filter(animal_type == "total_fish_bpue" | animal_type == "total_primary_fish_bpue" | animal_type == "halibut_bpue" | animal_type == "grouper_bpue" | animal_type == "total_secondary_fish_bpue") |> 
  pivot_wider(names_from = animal_type,
              values_from = bpue) |> 
  rename("total_fish_cpue" = "total_fish_bpue", 
         "total_primary_fish_cpue" = "total_primary_fish_bpue",
         "halibut_cpue" = "halibut_bpue",
         "grouper_cpue" = "grouper_bpue",
         "total_secondary_fish_cpue" = "total_secondary_fish_bpue") |> 
  pivot_longer(cols = c(total_fish_cpue, 
                        total_primary_fish_cpue, 
                        halibut_cpue, 
                        grouper_cpue, 
                        total_secondary_fish_cpue)) |> 
  rename("treatment"= "exp_unit",
         "catch" = "name",
         "cpue" = "value")
```

# `GLMM` on the BPUE data 
### Convert appropriate variables to be a factor
```{r}
bpue_glm <- bpue |> 
  mutate(catch = as.factor(animal_type), 
         treatment = as.factor(exp_unit), 
         gillnet_pair = as.factor(gillnet_pair))
```
### Initial exploration plot
```{r}
ggplot(bpue_glm, aes(x= gillnet_pair, y = bpue, color = catch)) + 
  geom_point() +
  facet_wrap(~treatment) + 
  theme_minimal()
```
## Fit the `gam()` for the bpue
```{r}
gam(bpue ~ treatment + s(gillnet_pair, bs = 're'),
    data = bpue_glm)

### Include tweedie distribution? 
# turtle_glmm <- gam(bpue ~ treatment + s(gill_net_pair, bs = "re"), 
#                    method = "REML", 
#                   family = tw(),
#                   # family = tweedie(link = "log"),
#                    data = turtle_tidy)
```
# Residual plots? 
# Confidence Intervals with bootstrapping 10,000 simulations
